Contextuality and Nonlocality in Discrete Quantum Theory

### 2016-01-26 DQC3.tex
Update "6.1 Gleason's Theorem in DQT"...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{amsfonts}
\usepackage{color}
\usepackage{stmaryrd}\usepackage{dsfont}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{enumitem}		% customizable list environments
\newlength{\lyxlabelwidth}      % auxiliary length 
\theoremstyle{plain}
\newtheorem{thm}{\protect\theoremname}
\ifx\proof\undefined
\newenvironment{proof}[1][\protect\proofname]{\par
\normalfont\topsep6\p@\@plus6\p@\relax
\trivlist
\itemindent\parindent
\item[\hskip\labelsep\scshape #1]\ignorespaces
}{%
\endtrivlist\@endpefalse
}
\providecommand{\proofname}{Proof}
\fi

@@

%%%%%%%%%%%%%%%%%
\subsection{Gleason's Theorem in DQT} The Born rule is the \emph{only}
way to assign a probability in conventional quantum theory because of
Gleason's theorem.  We now show that $\dpr $ defined in
Eq. (\ref{eq:modality}) is an example of a properly defined Born rule
in the discrete theory because $\dpr $ also satisfies the closest
analog of Gleason's Theorem that we can construct.

Following the standard axioms for Gleason's
theorem~\cite{peres1995quantum}, we introduce discrete axioms that
differ in only one significant instance from the classical ones:
\begin{eqnarray}
 & \bullet & \textrm{For any orthogonal basis }\left\{
  \left| n \right\rangle \right\}
  _{n=0}^{d-1}\subset\ffzd{p^{2}}\textrm{, and
  }\label{enu:total-probability-DQT(I)}\\ & &
  \ket{\Psi}\in\ffzd{p^{2}}\textrm{, at least one of\ }
  \dpr \left(\ket{\Psi},\ket{n}\right) = \text{possible.}\nonumber \\ &
  \bullet & \textrm{For any orthogonal basis }
\left\{  \left| n \right\rangle
\right\}_{n=0}^{d-1}\subset\ffzd{p^{2}}\nonumber \\ & &
 \textrm{and a particular\ }\ket{n'}\in\left\{ \left| n \right\rangle \right\}
  _{n=0}^{d-1}\textrm{, }\nonumber \\ & &
  \dpr\big(\ket{n'},\ket{n'}\big)=\text{possible and\ }
   \label{enu:particular-vector-DQT(I)}\\ & &
  \dpr\big(\ket{n'},\ket{n}\big)=\text{impossible for any }n\ne  n'
   \textrm{. }\nonumber \\ & \bullet & \textrm{For any vector  }
   \ket{\Psi},\ket{\Phi}\in\ffzd{p^{2}}
  \textrm{, and  }\gamma\in\ff{p^{2}}\backslash\left\{ 0\right\} 
  \textrm{, }\nonumber
  \\ & &
  \dpr (\ket{\Psi},\ket{\Phi})=\dpr (\gamma\ket{\Psi},\ket{\Phi}) = 
    \dpr (\ket{\Psi},\gamma\ket{\Phi}) \textrm{. }
  \label{enu:parallel-DQT(I)}\\
 & \bullet & \textrm{For any vector
  }\ket{\Psi},\ket{\Phi}\in\ffzd{p^{2}}
  \textrm{, }\nonumber \\ & &
  \textrm{and }U\textrm{ is a }d\textrm{-by-}d
  \textrm{ unitary matrix over }\ff{p^{2}}\textrm{, }
  \label{enu:unitary-DQT(I)}\\ & &
  \textrm{i.e. , }U^{\dagger}U=I\textrm{, then  }
   \dpr (\ket{\Psi},\ket{\Phi})=\dpr (U\ket{\Psi},U\ket{\Phi})\textrm{. }\nonumber
\end{eqnarray}
Because there is no possibility for continuity in the discrete theory,
we have substituted the condition (\ref{enu:unitary-DQT(I)}) for the
conventional continuity axiom in Gleason's Theorem.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{amsfonts}
\usepackage{color}
\usepackage{stmaryrd}\usepackage{dsfont}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{enumitem}		% customizable list environments
\newlength{\lyxlabelwidth}      % auxiliary length 
\theoremstyle{plain}
\newtheorem{thm}{\protect\theoremname}
\ifx\proof\undefined
\newenvironment{proof}[1][\protect\proofname]{\par
\normalfont\topsep6\p@\@plus6\p@\relax
\trivlist
\itemindent\parindent
\item[\hskip\labelsep\scshape #1]\ignorespaces
}{%
\endtrivlist\@endpefalse
}
\providecommand{\proofname}{Proof}
\fi
\theoremstyle{definition}
\newtheorem{problem}[thm]{Problem}

@@

%%%%%%%%%%%%%%%%%
\subsection{Gleason's Theorem in DQT}

Recall the measurement process discussed in section \ref{sec:Critical-Analysis-CQT}.
Given a physical system represented by a state $\ket{\Psi}$, and
an observable ${\mathbf{O}}$, whose eigenspectrum is given by ${\mathbf{O}}\left|i\right\rangle =\lambda_{i}\left|i\right\rangle $.
Then the Born rule of CQT tells us that the probability of obtaining
the eigenvalue $\lambda_{i}$ as an outcome of the measurement is
given by 
\begin{eqnarray}
\begin{array}{rcccl}
{\cal P}_{\Psi}\left(\ket{i}\right) & = & {\cal P}_{\Psi}\left(\lambda_{i}\right) & = & \frac{\ip{\Psi}{i}\ip{i}{\Psi}}{\ip{\Psi}{\Psi}\ip{i}{i}}\end{array}\ ,\label{prob1.eq-6}
\end{eqnarray}
If the system is a mixed state represented by a density matrix 
\[
\rho=\sum_{j}p_{j}\frac{\proj{\Psi_{j}}}{\ip{\Psi_{j}}{\Psi_{j}}}\textrm{ ,}
\]
we can extend the Born rule (\ref{prob1.eq-6}) and get 
\begin{eqnarray}
{\cal P}_{\rho}\left(\ket{i}\right)={\cal P}_{\rho}\left(\lambda_{i}\right) & = & \sum_{i}p_{i}\frac{\ip{\Psi_{j}}{i}\ip{i}{\Psi_{j}}}{\ip{\Psi_{j}}{\Psi_{j}}\ip{i}{i}}\label{prob1.eq-2}
\end{eqnarray}

Let $\C^{*}=\C\backslash\left\{ 0\right\} $ and $\C^{d*}=\left\{ \ket{\Psi}\in\C^{d}\mid\ip{\Psi}{\Psi}\ne0\right\} $
be the set of non-zero vectors. Because normalization and global phase
gives the same state, and ${\cal P}_{\rho}:\C^{d*}\rightarrow\left[0,1\right]$
should be a state-function, we should have 
\begin{equation}
\textrm{For any vector }\ket{i}\in\C^{d*}\textrm{, and }\gamma\in\C^{*}\textrm{ , }{\cal P}_{\rho}(\ket{i})={\cal P}_{\rho}(\gamma\ket{i})\textrm{ .}\label{eq:parallel-DQT(I)}
\end{equation}
More importantly, ${\cal P}_{\rho}$ defines a probability which means
\begin{equation}
\textrm{For any orthonormal basis }\{\ket{i}\}\textrm{ , }{\displaystyle \sum_{i=0}^{d-1}{\cal P}_{\rho}\left(\ket{i}\right)=1}\textrm{ .}\label{eq:total-probability}
\end{equation}
Conversely, if a state-function ${\cal P}:\C^{d*}\rightarrow\left[0,1\right]$
is a probability, i.e., satisfying equation (\ref{eq:parallel-DQT(I)})
and (\ref{eq:total-probability}), we can identify ${\cal P}$ to
a mixed state $\rho$ by Gleason's theorem.
\begin{thm}
(Gleason's Theorem) 
\begin{enumerate}
\item For $d\ge3$, for any state-function ${\cal P}:\C^{d*}\rightarrow\left[0,1\right]$,
if ${\cal P}$ is a probability, there is a density matrix $\rho$
such that ${\cal P}={\cal P}_{\rho}$~\cite{Redhead1987-REDINA,gleason1957}.\footnote{In order to simplify the statement, we consider consider the probability
on the state vectors instead of the projectors or closed subspaces
as in the Gleason's original paper.}
\item Consequently, the Born rule (\ref{prob1.eq-6}) is the only reasonable
probability assignment, and can be deduced by other reasonable axioms~\cite{peres1995quantum}. 
\end{enumerate}
\end{thm}
Notice that every state is an equivalence class of $\C^{d*}$, but
we cannot really get them directly by a physical measurement. All
we can get from a physical measurement is the probability ${\cal P}$.
An important meaning of Gleason's theorem is that the state space
is sufficient enough to produce every probability. Hence, we don't
have to introduce a new state to explain the probability got from
a measurement. This may be analog to the fact that every algebraic
equation over $\C$ can be solved in $\C$ so that we don't have to
introduce new numbers to solve an equation. The other important meaning
of Gleason's theorem is that it implies the Born rule (\ref{prob1.eq-6}).
This is even more important for building a discrete quantum theory.
If we plug in vectors over $\ff{p^{2}}$ into the usual Born rule
(\ref{prob1.eq-6}), it gives an elements in $\ff{p}$ which has no
way interpreted as a probability. Everyone who tried to build a quantum
theory over finite fields struggled on how to define a reasonable
probability. If Gleason's theorem works in any sense over finite fields,
we can have a reasonable way to induce the Born rule over finite fields.

A naive way to extend Gleason's Theorem in DQT may be as follow.
\begin{problem}
Given a state-function $\hat{\pi}:\ffzd{p^{2}}\rightarrow\mathscr{L}_{2}$
which is also a probability, i.e., 
\begin{eqnarray}
 &  & \textrm{For any vector }\ket{i}\in\ffzd{p^{2}}\textrm{, and }\gamma\in\ff{p^{2}}^{*}\textrm{ , }\hat{\pi}(\ket{i})=\hat{\pi}(\gamma\ket{i})\textrm{ . }\label{eq:parallel-DQT}\\
 &  & \textrm{For any orthonormal basis }\{\ket{i}\}\textrm{ , }{\displaystyle \bigvee_{i=0}^{d-1}\hat{\pi}\left(\ket{i}\right)=}\textrm{possible . }\label{eq:total-probability-DQT}
\end{eqnarray}
Can we say something about $\hat{\pi}$?
\end{problem}
The answer is no, and the reasons are simple. First, compare the following
arithmetic equation and Boolean equation.
\begin{eqnarray}
p_{1}+p_{2}+p_{3} & = & 1\label{eq:arithmetic}\\
\pi_{1}\vee\pi_{2}\vee\pi_{3} & = & \text{possible}\label{eq:Boolean}
\end{eqnarray}
When $p_{1}=\frac{1}{2}$, we still have an equality $p_{2}+p_{3}=1$
for $p_{2}$ and $p_{3}$ because $\left(\mathbb{R},0,+\right)$ is
a group. In contrast, when $\pi_{1}=\text{possible}$, we know nothing
about $\pi_{2}$ and $\pi_{3}$ because $\left(\mathscr{L}_{2},\text{impossible},\vee\right)$
is a semigroup without cancellative, i.e., $a\vee c=b\vee c$ does
not imply $a=b$. The difference between $\left(\mathbb{R},0,+\right)$
and $\left(\mathscr{L}_{2},\text{impossible},\vee\right)$ suggests
us to study the range of the probability. A reasonable range should
have the structure $\left(\mathscr{L},0,1,\oplus,<\right)$, where
$\left(\mathscr{L},0,\oplus,<\right)$ is the underlying commutative
ordered semigroup~\cite{clifford1958} because the condition (\ref{eq:total-probability})
and (\ref{eq:total-probability-DQT}) suggests the value of the range
should be able to add together. $1\in\mathscr{L}$ is another special
value to indicate the total probability. For example, $\left(\mathscr{L}_{2},\text{impossible},\text{possible},\vee,<\right)$
and $\left(\mathbb{R},0,1,+,<\right)$ are two possible range of the
probability, but we cannot pick $\mathscr{L}=\ff{p^{2}}$ because
we cannot define an inequality such that $\left(\ff{p^{2}},0,+,<\right)$
be an ordered semigroup.

Second, Gleason's theorem relies a lot on the continuous probability~\cite{Redhead1987-REDINA,gleason1957,peres1995quantum}.
However, continuity doesn't even make sense in DQT. This fact implies
that the structure of the domain of the probability, $\ffzd{p^{2}}$,
is also too weak. In order to think which structure we need, we can
check which the reasonable property ${\cal P}_{\Psi}:\C^{d*}\rightarrow\left[0,1\right]$
should have. Despite the total probability is one, ${\cal P}_{\Psi}$
must have 
\begin{eqnarray}
{\cal P}_{\Psi}\left(\ket{i}\right) & = & 0\textrm{ if }\ip{\Psi}{i}=0\label{prob1.eq-6-1}
\end{eqnarray}
where $\gamma\in\C^{*}$. Moreover, the probability must be invariant
under any rotation, i.e., 
\begin{equation}
{\cal P}_{U\Psi}\left(U\ket{i}\right)={\cal P}_{\Psi}\left(\ket{i}\right)\label{eq:rotation}
\end{equation}
for any unitary matrix $U$. Because we consider the probability of
two different state ${\cal P}_{\Psi}$ and ${\cal P}_{U\Psi}$ together,
it may be more reasonable to ask whether there is an unique ``Born
rule'' satisfying all the conditions, and our problem becomes
\begin{problem}
\label{prob:Gleason's}Is there a structure $\left(\mathscr{L},0,1,\oplus,<\right)$
such that $\tilde{\pi}:\ffzd{p^{2}}\rightarrow\left(\ffzd{p^{2}}\rightarrow\left\{ x\in\mathscr{L}\middle|0\le x\le1\right\} \right)$
satisfying the following conditions is unique for $d\ge3$ and infinitely
many for $d=2$?\footnote{Instead of equation (\ref{enu:particular-vector-DQT(I)}), another
reasonable condition is $\braket{\Psi}{i}=0\Leftrightarrow\tilde{\pi}_{\Psi}\left(\ket{i}\right)=0$.
However, when the underlying group $\mathscr{L}=\mathscr{L}_{2}$,
this condition implies an unique solution for $d=2$. This is too
strong and is not what we want. } 
\begin{eqnarray}
 & \bullet & \textrm{For any orthonormal basis }\left\{ \ket{i}\right\} _{n=0}^{d-1}\subset\ffzd{p^{2}}\textrm{, }\label{enu:total-probability-DQT(I)}\\
 &  & \textrm{ \ensuremath{{\displaystyle \bigoplus_{i=0}^{d-1}\tilde{\pi}_{\Psi}\left(\ket{i}\right)=1}} .}\nonumber \\
 & \bullet & \textrm{If }\braket{\Psi}{i}=0\textrm{ ,}\textrm{then }\tilde{\pi}_{\Psi}\left(\ket{i}\right)=0\textrm{ .}\label{enu:particular-vector-DQT(I)}\\
 & \bullet & \textrm{For any vector }\ket{\Psi},\ket{i}\in\ffzd{p^{2}}\textrm{, }\nonumber \\
 &  & \textrm{and }U\textrm{ is a }d\textrm{-by-}d\textrm{ unitary matrix over }\ff{p^{2}}\textrm{, }\label{enu:unitary-DQT(I)}\\
 &  & \textrm{i.e. , }U^{\dagger}U=I\textrm{, then }\tilde{\pi}_{\Psi}(\ket{i})=\tilde{\pi}_{U\Psi}(U\ket{i})\textrm{. }\nonumber \\
 & \bullet & \textrm{For any vector }\ket{\Psi},\ket{i}\in\ffzd{p^{2}}\textrm{, and }\gamma\in\ff{p^{2}}^{*}\textrm{, }\nonumber \\
 &  & \tilde{\pi}_{\Psi}(\ket{i})=\tilde{\pi}_{\Psi}(\gamma\ket{i})\textrm{. }\label{enu:parallel-DQT(I)}
\end{eqnarray}

\end{problem}
The no-go theorem in section \ref{sub:A-No-Go-Theorem} said the if
$\left(\mathscr{L},0,1,\oplus,<\right)=\left(\mathbb{R},0,1,+,<\right)$,
there is no $\tilde{\pi}:\ffzd{p^{2}}\rightarrow\ffzd{p^{2}}\rightarrow\mathscr{L}$
for $d\ge3$. It is easy to extend the no-go theorem and prove 
\begin{thm}
If $\left(\mathscr{L},0,\oplus,<\right)$ is a ordered group, then
there is no $\tilde{\pi}:\ffzd{p^{2}}\rightarrow\ffzd{p^{2}}\rightarrow\mathscr{L}$
satisfying all conditions in problem \ref{prob:Gleason's} for $d\ge3$.
\end{thm}
When we consider $\left(\mathscr{L}_{2},\text{impossible},\text{possible},\vee,<\right)$,
$\dpr$ defined in section \ref{sub:Interval-Valued-Probabilities-DQT}
satisfies all conditions, but $\dpr$ is not the unique solution because
of lack of cancellation law. It is still an open question whether
problem \ref{prob:Gleason's} is true or false in general.

\yutsung{Is that all probability value should be distinct be an reasonable
condition?}
